---
title: "Machine Learning assignment"
author: "Joris Schut"
date: "Tuesday, April 28, 2015"
output: html_document
---

##Initial steps
First, a initial check is performed to see if the files required are present. In not, the files are downloaded in the working directory. Second, the libraries required in the project are loaded.

##Loading the data
In this step, the data is loaded. This is done in two variables: one for the testing set and one for the training set.

###Creating the training and test sets
In this step the training data is split into two parts. 70% of the training data is put into the training set and the remaining 30% in the testing set (not to be confused with the testing data). 

##Cleaning the data
In order to run a random forest algorithm, the columns with more than 70% NA values were left out of the analysis. The columns that fulfilled this conditions in the training set were selected and their names were stored. In the testing set and the testing data the same columns were selected as in the training set.

##Fitting the model
Based on the remaining columns of the training set a random forest model (from the randomForest library) was fitted. This yielded a model with an accuracy of 100% on the training set.

```{r, comment=FALSE, echo=FALSE, message=FALSE, error=FALSE}
#Download the files (if not already done)
if (!file.exists("pml-training.csv")) {
  url  = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
  dest = "pml-training.csv"
  meth = "internal"
  quit = TRUE
  mode = "wb"
  download.file(url, dest, meth, quit, mode)
  #Works on tested operating system (Windows 7). Please change values if needed.
} 
if (!file.exists("pml-testing.csv")) {
  url  = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
  dest = "pml-testing.csv"
  meth = "internal"
  quit = TRUE
  mode = "wb"
  download.file(url, dest, meth, quit, mode)
  #Works on tested operating system (Windows 7). Please change values if needed.
} 

#Load libraries
library(dplyr)
library(caret)
library(randomForest)

#Load the training and testing data
trainingdata <- read.csv("pml-training.csv", na.strings=c("NA","#DIV/0!",""))
testingdata <- read.csv("pml-testing.csv", na.strings=c("NA","#DIV/0!",""))

#Divide the training set in a training and testing set
inTrain <- createDataPartition(y=trainingdata$classe,
                               p=0.7, list=FALSE)
training <- trainingdata[inTrain,]
testing <- trainingdata[-inTrain,]

#Cleaning and transformations

#Removing the x variable (# observation)
training <- training[c(-1)]
testing <- testing[c(-1)]
testingdata <- testingdata[c(-1)]

#Removing variables with many NA's (treshold: 70% NA)
training2 <- data.frame(matrix(nrow=nrow(training)))
trainingnames <- character()
a <- numeric()

#Training set
for(i in 1:length(training)){
  if(sum(is.na(training[1:nrow(training), i]))/nrow(training) <= 0.3){
    training2 <- cbind(training2, training[,i])
    trainingnames <- append(trainingnames, names(training[i]))
    a <- append(a,i)
  }
}

training2 <- training2[,-1]
names(training2) <- trainingnames
training <- training2

#Apply the same transformation to the test set
test2 <- data.frame(matrix(nrow=nrow(testing)))
for(j in 1:length(a)){
  test2 <- cbind(test2, testing[,a[j]])
}

test2 <- test2[,-1]
names(test2) <- trainingnames
testing <- test2

#Apply the same transformation to the final test set
test3 <- data.frame(matrix(nrow=nrow(testingdata)))
for(k in 1:length(a)){
  test3 <- cbind(test3, testingdata[,a[k]])
}

test3 <- test3[,-1]
names(test3) <- trainingnames
colnames(test3)[59] <- names(testingdata[159])
testingdata <- test3

#Develop the model

#Run the random forest algoritm on the training set
model <- randomForest(classe ~. , data=(training))

#Test the model

#Run the model on the testing set (derived from the training data)
predict <- predict(model, training, type="class")
predictions <- predict(model, testing, type = "class")

#Check the accuracy of the model
confusionMatrix(predict, training$classe)

````

##Predicting model accuracy
Model accuracy was predicted by fitting the data to the training set. Based on the testing set model accuracy was 99.9%,

```{r, echo=FALSE}
confusionMatrix(predictions, testing$classe)

```

##Placing the model over the test set
Finally, model was fitted on the testing data using the code provided by the instructors. Results were outputted in separate .txt files which were submitted to the Coursera website. Prediction was accurate in 20 out of 20 cases (100% accuracy).